Reduced Nucleotide Diversity on a Plant Y Chromosome Following Recent Recombination Suppression

### 2016-08-20 manuscript.tex
minor edits

To consider the joint effects of biased sex ratios and variance in male reproductive success, neutrally-expected levels of diversity were calculated using equations 1-3, assuming no variance in female mating (e.g., $C^2_{f}=0$), and we considered a range of values for $\sigma_{m}$: 0 (no variance), 1/2 (Poisson/2) 1 (Poisson distribution of offspring), 2 (Poisson*2), and 3 (Poisson*3). Figure 1 shows the predicted effective population size of autosomes and sex chromosomes as a function of the sex ratio under this model. Using these predictions, we calculated the expected X/A and Y/A ratios of diversity for the estimated \textit{R. hastatulus} sex ratio $N_{m}/N_{f}=0.6$ \citep{pickup2013influence}, assuming that the level of neutral polymorphism is given by the product of the mutation rate and the effective population size, $\theta=4N_{e}\mu$ \citep{watterson1975,kimura1984}. Note that we considered variance in male reproductive success here because this is most likely to occur in wind pollinated plant like \textit{R. hastatulus}, where males produce copious amounts of pollen and compete to fertilize uniovulate females (\X ref).

%old stuff
%The effect of biased reproductive sex ratios on the $N_{e}$ for an autosomal gene can be estimated using the equation \citep{wright1931evolution}:

%\begin{equation}
%N_{e} = \frac{4N_{m}N_{f}}{N_{m}+N_{f}}\label{eq:Ne}
%\end{equation}

%where $N_{m}$ and $N_{f}$ are the numbers of reproducing males and females, respectively. Similarly, for an X-linked gene, where females contain two-thirds and males one-third of the alleles, the effective population size is given by:

%\begin{equation}
%N_{e{X}} = \frac{9N_{m}N_{f}}{4N_{m}+2N_{f}}\label{eq:NeX}
%\end{equation}

%And, for the male-limited Y chromosome:

%\begin{equation}
%N_{e{Y}} = \frac{N_{m}}{2}\label{eq:NeY}
%\end{equation}

%\begin{equation}
%\frac{N_{e_{X}}}{N_{e_{A}}} = \frac{9(N_{f}+N_{m})}{8(2N_{f}+N_{m})}\label{eq:X/A}
%\end{equation}

%and,

%\begin{equation}
%\frac{N_{e_{Y}}}{N_{e_{A}}} = \frac{9(N_{f}+N_{m})}{8(2N_{f}+N_{m})}\label{eq:X/A}
%\end{equation}

 %Given a female effective population size $N_{f}$, and a male effective population size $N_{m}$, the predicted effective population size for X-linked genes, relative to autosomes, is given by $N_{e_{X}}/N_{e_{A}} = 9(N_{f}+N_{m})/8(2N_{f}+N_{m})$ (Wright 1931). Similarly, for the Y chromosome, the effective population size is given by 1/4 the total $N_{e}$ or 1/2 the $N_{e}$ of males, $N_{m}$/2. Thus, with equal sex ratios, $N_{e_{X}}/N_{e_{A}} = 0.75$, and $N_{e_{Y}}/N_{e_{A}} = 0.25$. For populations with female-biased sex ratios, however, $N_{m}$ becomes small relative to $N_{f}$ and this ratio can become larger than 1, approaching 1.125 at the limit \citep{caballero1995}. With the level of nucleotide polymorphism maintained in a population is proportional to the product of the mutation rate and the effective population size: $\theta=4N_{e}\mu$ \citep{watterson1975,kimura1984}. Figure 1 shows the predicted effective population size ratios for autosomal, X-linked, and Y-linked genes (left) and the corresponding X/A and Y/A ratios of diversity (right) for a sex ratio bias ranging from $N_{f}/(N_{f}+N_{m})=0$ to $N_{f}/(N_{f}+N_{m})=1$.

%In particular, Wright's (1931) formula $N_{e}=4N_{f}N_{m}/(N_{f} + N_{m})$ gives the effective population size of a randomly mating population with separate sexes and a Poisson distribution of offspring number. For a female effective population size $N_{f}$, and a male effective population size $N_{m}$, the predicted effective population size for X-linked genes relative to autosomes is given by $N_{e_{X}}/N_{e_{A}} = 9(N_{f}+N_{m})/8(2N_{f}+N_{m})$ (Wright 1931). With equal sex ratios, $N_{e_{X}}/N_{e_{A}} = 0.75$. For populations with female-biased sex ratios, however, $N_{m}$ becomes small relative to $N_{f}$ and this ratio can become larger than 1, approaching 1.125 at the limit \citep{caballero1995}. For the Y chromosome, the effective population size is given by 1/4 the total $N_{e}$ or 1/2 the $N_{e}$ of males, $N_{m}$/2. Under neutrality, the level of nucleotide polymorphism maintained in a population is proportional to the product of the mutation rate and the effective population size: $\theta=4N_{e}\mu$ \citep{watterson1975,kimura1984}. Assuming equal neutral mutation rates for all genes and an equal number of reproducing males and females, Y-linked genes should therefore have 1/3 the polymorphism of X-linked genes, and 1/4 that of autosomal genes. Figure 1 shows the predicted effective population size ratios for autosomal, X-linked, and Y-linked genes (left) and the corresponding X/A and Y/A ratios of diversity (right) for a sex ratio bias ranging from $N_{f}/(N_{f}+N_{m})=0$ to $N_{f}/(N_{f}+N_{m})=1$.

\begin{figure}[htbp]
\centering
\includegraphics[width=\linewidth]{Figure1.png}
\caption{The relation between relative effective population size and sex ratio bias for genes on autosomes (\textbf{A}) and sex chromosomes (\textbf{B}) (Y chromosome in red). The sex ratio is shown as the proportion of males, $N_{m}/(N_{f}+N_{m})$, where $N_{m}$ and $N_{f}$ are the effective number of breeding males and females, respectively, plotted against $N_{e}/N$, where $N=N_{m}+N_{f}$ and the $N_{e}$ for sex chromosomes and autosomes is given in equations 1-3. Dotted curves show predictions in the standard neutral model, where both males and females produce Poisson-distributed offspring numbers and the chromosomal $N_{e}$'s are given by equations 4-6 \citep{wright1931evolution}. Solid curves correspond to increasing levels of variance in male reproductive success \citep{nomura2002effective} (see Methods). Assuming $\theta=4N_{e}\mu$ and equal neutral mutation rates among genes, the predicted $N_{e}$'s are used to generate null predictions for X/A and Y/A ratios of diversity.  
}%
\label{fig:spectrum}
\end{figure}

\subsection*{Simulations of positive and purifying selection}
To test whether our observed level of Y-chromosome diversity could be explained by the effects of linked selection, we followed the approach used in \citep{Wilsonsayres2014} and conducted forward-time simulations using the software SFSCODE \citep{hernandez2008flexible} and compared our empirical diversity estimates with those from simulations. This approach was preferred over using analytical predictions primarily because the BGS model over-predicts the reduction in diversity when there are many linked sites under selection, which is expected to be the case for the Y chromosome where there is no crossing over. Moreover .....[\X explain here]

%We chose to use forward simulations, rather than using standard analytical background selection models, which assume the effect of background selection is a simple reduction in effective population size, for several reasons. First, the standard formulas were derived for equilibrium demographic models, but human populations have a more complex demographic history with unknown effects on the process of background selection.

%Second, many mutations have been shown to be weakly deleterious and may persist in the population due to genetic drift [37,38]. The standard theory does not allow for this. Finally, simulations studies suggest that the standard theory can over-predict the reduction in genetic diversity due to background selection if there are many weakly selected linked mutations [26]. The forward simulations that we conducted address all of these concerns.

Because we were interested primarily in contrasting the

(i) Purifying selection acting to remove deleterious mutations and linked neutral variation (i.e., background selection), and (ii) Positive selection acting on beneficial mutations and reducing variation through selective sweeps (refs \X). For purifying selection simulations, we assumed that selection coefficients followed a gamma distribution, and we considered a range of values for the mean selection coefficient against deleterious mutations ($s$) by varying the scale parameter, $k$ from \X to \X while keeping the distribution shape parameter constant (see Supporting Information for simulation commands). To make the simulation output comparable to our data, we initialized the simulations with our estimated value of $\theta$ (estimated from autosomes), and sampled 8 chromosomes per simulation from 24 populations.

For simulating positive selection on the Y, we included a neutral locus of length Ln = 103 bp that is flanked by loci on either side that are subject to selection and are each of length Lf = 105 bp.

In a forward simulation, you can only specify the mutation rate, not the substitution rate. So, for a given substitution rate  , we must divide by the fixation probability to obtain the corresponding mutation rate of selected mutations. Let ⌫ be the per base pair mutation rate in the flanking loci:

Simulated sequences also contained 500 kb of linked neutral sequence from which we calculated diversity to asses the effects of selection in each model. We ran 1000 replicates for each model and calculated the average $\theta_{sim}$ and the corresponding Y/A ratio of neutral diversity under the both sets of models separately. We then tested the significant of the fit of our simulated diversity output to our empirical data (\X new figure) by calculating the proportion of simulation replicates for which $\theta_{sim}$ was not significantly different from $\theta_{observed}$, using a two-sided P value. Commands for running the simulations are provided in the Supporting Information.

>>>>

To consider the joint effects of biased sex ratios and variance in male reproductive success, neutrally-expected levels of diversity were calculated using equations 1-3, assuming no variance in female mating (e.g., $C^2_{f}=0$), and we considered a range of values for $\sigma_{m}$: 0 (no variance), 1/2 (Poisson/2) 1 (Poisson distribution of offspring), 2 (Poisson*2), and 3 (Poisson*3). Figure 1 shows the predicted effective population size of autosomes and sex chromosomes as a function of the sex ratio under this model. Using these predictions, we calculated the expected X/A and Y/A ratios of diversity for the estimated \textit{R. hastatulus} sex ratio $N_{m}/N_{f}=0.6$ \citep{pickup2013influence}, assuming that the level of neutral polymorphism is given by the product of the mutation rate and the effective population size, $\theta=4N_{e}\mu$ \citep{watterson1975}. Note that we considered variance in male reproductive success here because this is most likely to occur in wind pollinated plant like \textit{R. hastatulus}, where males produce copious amounts of pollen and compete to fertilize uniovulate females (\X ref).

\begin{figure}[htbp]
\centering
\includegraphics[width=\linewidth]{Figure1.png}
\caption{The relation between relative effective population size and sex ratio bias for genes on autosomes (\textbf{A}) and sex chromosomes (\textbf{B}) (Y chromosome in red). The sex ratio is shown as the proportion of males, $N_{m}/(N_{f}+N_{m})$, where $N_{m}$ and $N_{f}$ are the effective number of breeding males and females, respectively, plotted against $N_{e}/N$, where $N=N_{m}+N_{f}$ and the $N_{e}$ for sex chromosomes and autosomes is given in equations 1-3. Dotted curves show predictions in the standard neutral model, where both males and females produce Poisson-distributed offspring numbers and the chromosomal $N_{e}$'s are given by equations 4-6 \citep{wright1931evolution}. Solid curves correspond to increasing levels of variance in male reproductive success \citep{nomura2002effective} (see Methods). Assuming $\theta=4N_{e}\mu$ and equal neutral mutation rates among genes, the predicted $N_{e}$'s are used to generate null predictions for X/A and Y/A ratios of diversity.  
}
\label{fig:spectrum}
\end{figure}

\subsection*{Simulations of positive and purifying selection}
To test whether our observed level of Y-chromosome diversity could be explained by the effects of linked selection, we followed the approach used in \citep{Wilsonsayres2014} and conducted forward-time simulations using the software SFSCODE \citep{hernandez2008flexible} and compared our empirical diversity estimates with those from simulations. This approach was preferred over using analytical predictions primarily because the background selection model over-predicts the reduction in diversity when there are many linked sites under selection \citep{KaiserCharlesworth}, as is expected to be the case for Y chromosomes that lack crossing over.

For purifying selection simulations, we assumed that selection coefficients followed a gamma distribution, and we considered a range of values for the mean selection coefficient against deleterious mutations ($s$) by varying the scale parameter, $k$ from \X to \X while keeping the distribution shape parameter constant (see Supporting Information for simulation commands). To make the simulation output comparable to our data, we initialized the simulations with our estimated value of $\theta$ (estimated from autosomes), and sampled 8 chromosomes per simulation from 24 populations. For simulating positive selection on the Y, we included a neutral locus of length Ln = 103 bp that is flanked by loci on either side that are subject to selection and are each of length Lf = 105 bp. Simulated sequences also contained 500 kb of linked neutral sequence from which we calculated diversity for each model. We ran 1000 replicates for each model and calculated the average $\theta_{sim}$ and the corresponding Y/A ratio of neutral diversity under the both sets of selection models separately. Finally, we tested the significance of the fit of our simulated diversity output to our empirical data (\X new figure) by calculating the proportion of simulation replicates for which $\theta_{sim}$ was not significantly different from $\theta_{observed}$, using a two-sided P value. Commands for running the simulations are provided in the Supporting Information.
